The Effect of Episodic Retrieval on Inhibition in Task Switching

### 2015-11-27 2015/Episodic Retrieval & Inhibition/Manuscript Files/main.tex
minor update to manuscript file

\abstract{Inhibition in task switching is inferred from n--2 task repetition costs: the observation that AB$A$ task switching sequences are responded to slower than CB$A$ sequences. This is thought to reflect the persisting inhibition of task A, which slows re-activation attempts. \cite{Mayr2002} reported an experiment testing a critical non-inhibitory account of this effect, namely episodic retrieval: If the trial parameters for task A match across an ABA sequence, responses should be facilitated due to priming from episodic retrieval; a cost would occur if trial parameters mismatch. In a rule-switching paradigm, Mayr reported no significant difference in n--2 repetition cost when the trial parameters repeated or switched across an ABA sequence, in contrast to the episodic retrieval account. Here we provide a replication of the critical aspects of Mayr's design. Using Bayesian analysis to quantify evidence for or against a null hypothesis, we find evidence for reduced n--2 repetition costs when trial parameters repeat compared to when they switch. In contrast to Mayr's conclusion, we suggest that episodic retrieval can influence the measure of inhibition in task switching, although it cannot explain it entirely.}

@@

How the cognitive system solves the stability--flexibility dilemma has been studied using the task switching paradigm \citep{Grange2014a,Kiesel2010,Vandierendonck2010}, wherein participants have to make speeded responses to simple cognitive tasks. For example, participants might be presented with number stimuli, and be asked to switch between judging whether the stimulus is odd/even or lower/higher than five. One cognitive process thought to be essential for task switching performance is the inhibition of competing tasks. Evidence for inhibition comes from the so-called backward inhibition paradigm \citep{Koch2010, Mayr2000}, where participants are required to switch between three tasks (A, B, and C).  It is a consistent finding that AB$A$ task switching sequences are performed slower and with less accuracy than CB$A$ sequences. This \emph{n--2 task repetition cost} is thought to reflect the persisting inhibition of task A in an ABA sequence, which hinders its re-activation in the last trial of the triplet.

The n--2 task repetition cost is important as it appears to be robust against non-inhibitory explanations \citep{Mayr2007}. One potential explanation of the n--2 task repetition cost that does not assume inhibition---the episodic retrieval account \citep{Neill1997}---was explored by \cite{Mayr2002}. This theory suggests (adapted to a task switching context) that when a task has been performed an episodic trace of trial parameters (such as the cue, stimulus characteristics, and the response made) is stored in memory. When this task is cued again, retrieval of the most recent episodic trace occurs. If the parameters of the presented trial differ from that of the retrieved episode (e.g., a different response is required), a mismatch cost occurs relative to if the retrieved episode matches parameters of the current trial (which will prime performance). Thus, according to this account, n--2 task repetition costs reflect a mismatch cost, as trial parameters typically differ between both instances of task A in an ABA sequence.

\cite{Mayr2002} assessed this alternative explanation by developing a paradigm where the match between trial parameters across an ABA sequence could be manipulated (see Figure \ref{fig:mayrExperiment}). Participants were required to apply a spatial transformation of a target's location according to one of three stimulus--response rules (vertical, horizontal, or diagonal transformations). For example, if the target is in the bottom-left of the display and the current rule is ``vertical'', the correct response would be an upper-left response (as this reflects a vertical transformation of a bottom-left target). The parameters that might be stored in an episodic trace include (at least) the cue presented, the target's location, and the response made. In this paradigm, trial parameters can either match or mismatch across ABA and CBA sequences\footnote{Note that with one cue per task, the cue for task A will always be the same across an ABA sequence.}. For example, if the target is in the same location for task A across an ABA sequence, the same response will be required (i.e., an n--2 response repetition). In this case, the episodic retrieval account would predict facilitated performance relative to an n--2 response switch as in the former case the current trial parameters match perfectly the retrieved episodic trace. Thus, the episodic account would predict reduced (or zero) n--2 task repetition costs for n--2 response repetitions compared to n--2 response switches (where there is a mismatch between trial parameters and the retrieved episodic trace).

@@

\section{The Current Study}
The purpose of the current study was to re-examine the evidence for the effect of episodic retrieval on n--2 task repetition costs. The conclusion of \cite{Mayr2002} rests on the acceptance of a null hypothesis (i.e., no difference of n--2 task repetition costs), which cannot be achieved via standard null hypothesis significance testing \citep[NHST;][]{Gallistel2009, Wagenmakers2007}; however, evidence in favour of a null hypothesis (given some data) can be obtained using Bayes Factors \citep[e.g.,][]{Rouder2009}. The Bayes Factor (denoted $BF_{10}$) quantifies the relative evidence provided by the data in favour of an alternative hypothesis (i.e., n--2 task repetition costs are different for response repetitions and response switches) over a null hypothesis (i.e., n--2 repetition costs are equivalent).

The Bayes factor for the critical statistical test from Mayr's (2002) study ($BF_{10}$ = 0.315)\footnote{For all Bayes factors reported in this paper we use a non-informative prior on the alternative hypothesis, which is assumed to be distributed as a Cauchy with scaling factor $r$ = 0.707.} suggests that the null hypothesis is about 3 times more likely than the alternative, given the data, which provides anecdotal-to-moderate support for the null. As this Bayesian analysis does not provide decisive evidence for the null, and given the importance of assessing to what extent (if any) episodic retrieval can influence n--2 task repetition costs in task switching, we were interested in performing a replication of Mayr's study to gather more evidence.

The current study replicated critical aspects of Mayr's (2002) design relevant to n--2 task repetitions. As Mayr's study was also interested in exploring to what extent the rule-switching paradigm generated typical task switching effects (such as switch costs, effects of preparation, and effects of the delay between tasks), we made some changes to the original design of Mayr. Specifically, immediate task repetitions were not allowed; it has been shown that including immediate task repetitions reduces the n--2 task repetition cost \citep{Philipp2006}, so we removed this possibility. We also did not manipulate preparation via the cue--stimulus interval, as Mayr reported no effect of this variable on n--2 task repetition costs. Mayr did find that n--2 repetition costs were larger with shorter response--cue intervals (RCI; the time between the response to one trial and the onset of the cue for the next trial), a typical finding for n--2 task repetition costs \citep{Gade2005, Grange2009, Mayr2000}. Therefore, in order to obtain large n--2 task repetition costs we only used a short RCI.

@@

\subsection{Apparatus \& Stimuli}
Stimuli were presented on a 17in. monitor connected to a PC running E-Prime v. 2.0 software. The stimulus display consisted of an 8cm by 8cm square frame. A black circle (diameter = 1cm) served as the trial stimulus. Possible cues were the words ``horizontal'', ``vertical'', or ``diagonal'', and were presented in size 22 black Verdanna font directly above the stimulus frame. Responses were collected via a 1-ms precise USB keyboard.

@@

\subsection{Design}
The experiment manipulated two factors in a fully-related design: \emph{Task Sequence} (n--2 task repetition [ABA] vs. n--2 task switch [CBA]) and \emph{Response Repetition} (n--2 response repetition vs. n-- response switch). The dependent variables were response time (RT) in milliseconds (ms) and percent accuracy (\%).

@@

\begin{figure*}
\begin{center}
\includegraphics[width =\textwidth]{Images/recruitmentOrder.pdf}
\caption{Simulated progression of the Sequential Bayes Factors for a random set of 50 different recruitment orders for the current data set. See text for details.}
\label{fig:recruitmentOrder}
\end{center}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width =\textwidth]{Images/bayesParameter.pdf}
\caption{Bayesian posterior distributions for the parameters of mean difference and effect size of difference for n--2 task repetition cost [response repetition] minus n--2 task repetition cost [response switch]. HDI = Highest Density Interval. See text for details.}
\label{fig:bayesParameter}
\end{center}
\end{figure*}

These posterior distributions present histograms of the Bayesian posterior estimate of the true mean and effect size underlying the data we obtained. The black horizontal bar represents the 95\% Highest Density Interval (HDI); in contrast to frequentist confidence intervals, we can interpret this interval as the span of parameter values we are 95\% certain the true underlying parameter lies within. The mean difference in n--2 task repetition costs for response repetitions and response switches is estimated to be -37ms; the HDI does not cross zero, so we can be confident that there is a difference between response repetitions and switches. Note, though, that the effect size of this difference is in the small range. For hypothesis testing (i.e., assessing whether episodic retrieval does impact the n--2 repetition cost or not), this is less relevant, but it is important to note if we wish to explore \emph{how much} episodic retrieval might affect n--2 repetition costs.

>>>>

\abstract{Inhibition in task switching is inferred from n--2 repetition costs: the observation that AB$A$ task switching sequences are responded to slower than CB$A$ sequences. This is thought to reflect the persisting inhibition of task A, which slows re-activation attempts. \cite{Mayr2002} reported an experiment testing a critical non-inhibitory account of this effect, namely episodic retrieval: If the trial parameters for task A match across an ABA sequence, responses should be facilitated due to priming from episodic retrieval; a cost would occur if trial parameters mismatch. In a rule-switching paradigm, Mayr reported no significant difference in n--2 repetition cost when the trial parameters repeated or switched across an ABA sequence, in contrast to the episodic retrieval account. This conclusion rests on accepting a null hypothesis, which cannot be achieved with null hypothesis significance testing. Here we provide a replication of the critical aspects of Mayr's design. Using Bayesian analysis to quantify evidence for or against a null hypothesis, we find evidence for reduced n--2 repetition costs when trial parameters repeat compared to when they switch. In contrast to Mayr's conclusion, we suggest that episodic retrieval can influence the measure of inhibition in task switching, although it cannot explain it entirely.}

@@

How the cognitive system solves the stability--flexibility dilemma has been studied using the task switching paradigm \citep{Grange2014a,Kiesel2010,Vandierendonck2010}, wherein participants have to make speeded responses to simple cognitive tasks. For example, participants might be presented with number stimuli, and be asked to switch between judging whether the stimulus is odd/even or lower/higher than five. One cognitive process thought to be essential for task switching performance is the inhibition of competing tasks. Evidence for inhibition comes from the backward inhibition paradigm \citep{Koch2010, Mayr2000}, where participants are required to switch between three tasks (A, B, and C).  It is a consistent finding that AB$A$ task switching sequences are performed slower and with less accuracy than CB$A$ sequences. This \emph{n--2 task repetition cost} is thought to reflect the persisting inhibition of task A in an ABA sequence, which hinders its re-activation in the last trial of the triplet.

The n--2 task repetition cost is important as it appears to be robust against non-inhibitory explanations \citep{Mayr2007}. One potential explanation of the n--2 task repetition cost that does not assume inhibition was explored by \cite{Mayr2002}. This \emph{episodic retrieval} account \citep{Neill1997} suggests (adapted to a task switching context) that when a task has been performed an episodic trace of trial parameters (such as the cue, stimulus characteristics, and the response made) is stored in memory. When this task is cued again, retrieval of the most recent episodic trace occurs. If the parameters of the presented trial differ from that of the retrieved episode (e.g., a different response is required), a mismatch cost occurs relative to if the retrieved episode matches parameters of the current trial (which will prime performance). Thus, according to this account, n--2 task repetition costs reflect a mismatch cost, as trial parameters typically differ between both instances of task A in an ABA sequence.

\cite{Mayr2002} assessed this alternative explanation by developing a paradigm where the match between trial parameters across an ABA sequence could be manipulated (see Figure \ref{fig:mayrExperiment}). Participants were required to apply a spatial transformation of a target's location according to one of three stimulus--response rules (vertical, horizontal, or diagonal transformations). For example, if the target is in the bottom-left of the display and the current rule is ``vertical'', the correct response would be an upper-left response (as this reflects a vertical transformation of a bottom-left target). The parameters that might be stored in an episodic trace include (at least) the cue presented, the target's location, and the response made. In this paradigm, trial parameters can either match or mismatch across ABA and CBA sequences\footnote{Note that with one cue per task, the cue for task A will always be the same across an ABA sequence.}. For example, if the target is in the same location for task A across an ABA sequence, the same response will be required (i.e., an n--2 response repetition). In this case, the episodic retrieval account would predict facilitated performance relative to an n--2 response switch as in the former case the current trial parameters match perfectly the retrieved episodic trace. Thus, the episodic account would predict reduced n--2 task repetition costs for n--2 response repetitions compared to n--2 response switches (where there is a mismatch between trial parameters and the retrieved episodic trace).

@@

\section{The Current Study}
The purpose of the current study was to re-examine the evidence for the effect of episodic retrieval on n--2 task repetition costs. The conclusion of \cite{Mayr2002} rests on the acceptance of a null hypothesis (i.e., no difference of n--2 task repetition costs), which cannot be achieved via standard null hypothesis significance testing \citep{Gallistel2009, Wagenmakers2007}; however, evidence in favour of a null hypothesis (given some data) can be obtained using Bayes factors \citep[e.g.,][]{Rouder2009}. The Bayes factor (denoted $BF_{10}$) quantifies the relative evidence provided by the data in favour of an alternative hypothesis (i.e., n--2 task repetition costs are different for response repetitions and response switches) over a null hypothesis (i.e., n--2 repetition costs are equivalent).

The Bayes factor for the critical statistical test from Mayr's (2002) study ($BF_{10}$ = 0.315)\footnote{For all Bayes factors reported in this paper we use a non-informative prior on the alternative hypothesis, which is assumed to be distributed as a Cauchy with scaling factor $r$ = 0.707.} suggests that the null hypothesis is about 3 times more likely than the alternative, given the data, which provides anecdotal-to-moderate support for the null \citep[e.g.,][]{Schoenbrodtinpress}. As this Bayesian analysis does not provide compelling evidence for the null, and given the importance of assessing to what extent (if any) episodic retrieval can influence n--2 task repetition costs in task switching, we were interested in performing a replication of Mayr's study to gather more evidence.

The current study replicated critical aspects of Mayr's (2002) design relevant to n--2 task repetitions. As Mayr's study was also interested in exploring to what extent the rule-switching paradigm generated typical task switching effects (such as switch costs, effects of preparation, and effects of the delay between tasks), we made some changes to the original design so that it was more tailored toward assessing n--2 task repetition costs. Specifically, immediate task repetitions were not allowed; it has been shown that including immediate task repetitions reduces the n--2 task repetition cost \citep{Philipp2006}, so we removed this possibility. We also did not manipulate preparation via the cue--stimulus interval, as Mayr reported no effect of this variable on n--2 task repetition costs. Mayr did find that n--2 repetition costs were larger with shorter response--cue intervals (RCI; the time between the response to one trial and the onset of the cue for the next trial), a typical finding for n--2 task repetition costs \citep{Gade2005, Grange2009, Mayr2000}. Therefore, in order to obtain larger n--2 task repetition costs we only used a short RCI.

@@

\subsection{Apparatus \& Stimuli}
Stimuli were presented on a 17in. monitor connected to a PC running E-Prime v. 2.0 software. The stimulus display consisted of an 8cm by 8cm square frame. A black circle (diameter = 1 cm.) served as the trial stimulus. Possible cues were the words ``horizontal'', ``vertical'', or ``diagonal'', and were presented in size 22 black Verdanna font directly above the stimulus frame. Responses were collected via a 1-ms. precise USB keyboard.

@@

\subsection{Design}
The experiment manipulated two factors in a fully-related design: \emph{Task Sequence} (n--2 task repetition [ABA] vs. n--2 task switch [CBA]) and \emph{Response Repetition} (n--2 response repetition vs. n-- response switch). The dependent variables were response time (RT) in milliseconds (ms.) and percent accuracy (\%).

@@

\begin{figure*}
\begin{center}
\includegraphics[width =  \textwidth]{Images/recruitmentOrder.pdf}
\caption{Simulated progression of the Sequential Bayes Factors for a random set of 50 different recruitment orders for the current data set. See text for details.}
\label{fig:recruitmentOrder}
\end{center}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width = \textwidth]{Images/bayesParameter.pdf}
\caption{Bayesian posterior distributions for the parameters of mean difference and effect size of difference for n--2 task repetition cost [response repetition] minus n--2 task repetition cost [response switch]. HDI = Highest Density Interval. See text for details.}
\label{fig:bayesParameter}
\end{center}
\end{figure*}

These posterior distributions present histograms of the Bayesian posterior estimate of the true mean and effect size underlying the data we obtained. The black horizontal bar represents the 95\% Highest Density Interval (HDI); in contrast to frequentist confidence intervals, we can interpret this interval as the range of parameter values we are 95\% certain the true underlying parameter lies within. The mean difference in n--2 task repetition costs for response repetitions and response switches is estimated to be -37ms; the HDI does not cross zero, so we can be confident that there is a difference between response repetitions and switches. Note, though, that the effect size of this difference is in the small range. For hypothesis testing (i.e., assessing whether episodic retrieval does impact the n--2 repetition cost or not), this is less relevant, but it is important to note if we wish to explore \emph{how much} episodic retrieval might affect n--2 repetition costs.
